---
title: "Titanic Data Sets and RandomForests"
author: "Trey Beeman"
date: "December 9, 2016"
output: html_document
---
# __1 Introduction__

### __1.1 Abstract__

This document is being created to learn `randomForests` working with data from Kaggle competitions.  Thanks to Kaggle.com and Datacamp.com for the information and guidance in preparing this document.

### __1.2 Libraries__
This is an exhaustive list of all libraries used for this project.
```{r libraries, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(randomForest)
library(rpart)
library(caret)
library(rattle)
library(rpart.plot)
library(RColorBrewer)
```

# __2 Data__
We read the data sets into R.  The data can be found at the links in the code block. 
```{r load datasets}
# create url strings
train_url <- "http://s3.amazonaws.com/assets.datacamp.com/course/Kaggle/train.csv"
test_url <- "http://s3.amazonaws.com/assets.datacamp.com/course/Kaggle/test.csv"

# read the data in to R.
train <- read.csv(train_url, stringsAsFactors = FALSE)
test <- read.csv(test_url, stringsAsFactors = FALSE)

# combine the data into a single dataframe for feature engineering
df <- bind_rows(train,test)

```

### __2.1 Exploratory Data Analysis__ 

We take a look at the data to gain insight into what can be pulled out of it.

A `str()` of the data with an overview of the data, class and variabe names. 
```{r, echo=FALSE}
str(df)
```

We will highlight a few useful or interesting facts from the data.

This is a proportion table of the survivors by `Sex`.
```{r, echo=FALSE}
prop.table(table(df$Sex, df$Survived),1)
```

This is a proportion table of survivors by Passenger Class `Pclass`.
```{r, echo=FALSE}
prop.table(table(df$Pclass, df$Survived),1)
```

##### __2.1.1 The Passengers__
The mean age of all passengers is: 

```{r, echo = FALSE}
mean(df$Age,na.rm = TRUE)
```

The mininum age of all passengers is:
```{r, echo = FALSE}
min(df$Age,na.rm = TRUE)
```

The maxinum age of all passengers is:
```{r, echo = FALSE}
max(df$Age,na.rm = TRUE)
```

##### __2.2.2 The Survivors__
The mean age of survivors is:
```{r, echo = FALSE}
mean(df$Age[df$Survived == 1],na.rm = TRUE)
```

The mininum age of survivors:
```{r, echo = FALSE}
min(df$Age[df$Survived == 1],na.rm = TRUE)
```

The maximum age of survivors:
```{r, echo = FALSE}
max(df$Age[df$Survived == 1],na.rm = TRUE)
```

### __2.2 Exploratory Graphs__

```{r exploratory graphs, echo = FALSE}
hist(df$Age, breaks = seq(0, 80, 5), main = "Histogram of Ages of All Passengers", xlab = "Passenger Ages in groups of 5 years")
hist(df$Age[train$Survived == 1], breaks = seq(0, 80, 5), main = "Histogram of Ages of Surviving Passengers", xlab = "Survivors Ages in groups of 5 years")
```

# __3 Feature Engineering__

### __3.1 Names__
We will create or __feature engineer__ a new variable based first on names.  We create a `Title` variable.
```{r titles}
# pull the title out of the passenger names
df$Title <- gsub('(.*, )|(\\..*)', '', df$Name)

table(df$Sex, df$Title)
```

First, we will quickly clean up the titles with are synonyms and congnates.
```{r titles cognates}
df$Title[df$Title == "Miss"] <- "Ms"
df$Title[df$Title == "Mlle"] <- 'Ms'
df$Title[df$Title == "Mme"] <- 'Mrs'
df$Title[df$Title == "Master"] <- 'Mr'
```

With the remaining, we will divide them into classes of __nobility, military, Dr__ and __Rev__ in order to preserve the information in these titles.

```{r titles honorific}
nobility <- c("Don", "Dona", "Jonkheer", "Lady", "Sir", "the Countess")
military <- c("Capt", "Col", "Major")

df$Title[df$Title %in% nobility] <- "Nobility"
df$Title[df$Title %in% military] <- "Military"
```

It is interesting to note that all the individuals with the title __Rev__ did not survive.
```{r titles graph, echo=FALSE}
ggplot(data = df[(df$Title %in% c("Dr", "Military", "Nobility", "Rev")) & (!is.na(df$Survived)),], aes(x = Title, fill = factor(Survived))) + geom_bar(stat='count', position='dodge') + labs(x = 'Titles')
```

### __3.2 Families__ 

We want to examine the hypothesis that families are likely to move together as a group, therefore, their survivorship may differ from an individual travelling alone.  We will create `family_size` variable to indicate a family travelling together.

```{r families 1}
# create a new variable, family_size
df$family_size <- df$SibSp + df$Parch + 1

# Create a family variable identifying a group travelling together
df$Family <- paste(df$Surname, df$family_size, sep='_')
```

We can see that individuals and familes greater than 4 had a lower survivorship than other groups. 
```{r families graph 1}
ggplot(data = subset(df, !is.na(df$Survived)), aes(x = family_size, fill = factor(Survived))) + geom_bar(stat='count', position='dodge') + scale_x_continuous(breaks=c(1:11)) + labs(x = 'Family Size')
```

We are going to created a __discretized__ variable on family size in order to examine this data a little more closely.

```{r discetize family var}
# create the vatiables
df$Disfamily_size[df$family_size == 1] <- 'singleton'
df$Disfamily_size[df$family_size < 5 & df$family_size > 1] <- 'small'
df$Disfamily_size[df$family_size > 4] <- 'large'

# mosaicplot of the 
mosaicplot(table(df$Disfamily_size, df$Survived), main='Family Size by Survival', shade=TRUE)
```

The mosaic plot confirms our hyopthesis that there is a penalty for individuals and large families.

### __3.3 Cabin__



# __4 Missing Values__

In order to continue this analysis, we must address the misssing values.


# __5 Modelling__

We will begin by looking at a decision tree for important clusters of data
```{r decision tree}
D.tree <- rpart(Survived ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked, data = train, method = "class")
fancyRpartPlot(D.tree)
```


set.seed(111)

# Apply the Random Forest Algorithm
my_forest <- randomForest(as.factor(Survived) ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Title, data = train, importance = TRUE, ntree = 1000)

# Make your prediction using the test set
my_prediction <- predict(my_forest, test)

# Create a data frame with two columns: PassengerId & Survived. Survived contains your predictions
my_solution <- data.frame(PassengerId = test[,1], Survived = my_prediction)

# Write your solution away to a csv file with the name my_solution.csv
write.csv(my_solution, file = "my_solution.csv", row.names = FALSE)

varImpPlot(my_forest)
